---
title: "Generate Scope 3 disaggregation table (Tier 1/2/3+)"
author: "Damien Lieber @ DecarbNexus LLC"
date: "`r Sys.Date()`"
# No output format on purpose: run via scripts/run_analysis.R or Run All Chunks
---
<!--
BEGINNER NOTES
- You do not need to edit this file to use the outputs. Most users can just
  download the Excel/CSVs in the outputs/ folder on GitHub.
- To reproduce the outputs locally: run `source("scripts/run_analysis.R")` in R,
  or open this file and use "Run All Chunks" (no Pandoc needed).
- This document will:
  1) Install/load required R packages as needed
  2) Download the correct USEEIO model spec based on config.yml
  3) Build the model and compute Tier/Scope contributions
  4) Write Excel and CSV to outputs/
-->

```{r setup, include=FALSE}
# Global chunk options: hide code by default to keep the document readable.
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
# Normalize working directory: when knitting directly from RStudio the default
# working directory is the Rmd's folder (analysis/); reset to project root so
# all relative paths (config.yml, outputs/, spec_files/, etc.) resolve correctly.
if (basename(getwd()) == "analysis") setwd("..")
```

## 0. Read this first (for beginners)

- You can run this from RStudio or VS Code with: `source("scripts/run_analysis.R")`.
- The results will be saved to the `outputs/` folder.
- Configuration lives in `config.yml` (SEF version and which sectors count as Scope 2).

## 1. Introduction

This R Markdown document performs a Structural Path Decomposition (SPD) on a specified USEEIO model. The goal is to disaggregate the total supply chain emission factors for each commodity into:

* **Tiers:** The position in the supply chain (Economic Tier 1 = the commodity itself, Economic Tier 2 = direct suppliers, Economic Tier 3+ = all upstream suppliers).
* **Scopes:** The GHG accounting category of the emissions (Scope 1 or Scope 2).

The final outputs are an Excel workbook and a CSV, saved in the `outputs/` folder.

## 2. Setup & Configuration

This block handles all setup, including package installation and dynamically fetching model specifications from the official USEPA repository based on the user's selected `sef_version`.

```{r 01-setup-and-load-packages, echo=FALSE, message=FALSE, warning=FALSE}
# --- 1. Package Management ---
if (!require("pacman")) install.packages("pacman")
pacman::p_load(
  yaml,       # For reading configuration files (local and remote)
  dplyr,      # For data manipulation
  reshape2,   # For the melt() function
  knitr,      # For creating nice tables
  httr,       # For checking URLs before trying to read them
  openxlsx,   # For writing to Excel files
  stringr,    # For advanced string manipulation
  tidyr,      # For separate() function
  cli,        # For concise, readable status messages
  arrow,      # For Parquet export (optimized columnar format)
  jsonlite    # For hierarchical JSON and JSON-LD export
)

# --- 2. Load User Configuration ---
config <- read_yaml("config.yml")
cat("User requested Supply Chain Factors (SEF) version:", config$sef_version, "\n")

# --- 3. Fetch Model Specifications from USEPA GitHub Repository ---
versioning_url <- "https://raw.githubusercontent.com/USEPA/supply-chain-factors/main/Versioning.yml"

# Check if the URL is accessible
response <- GET(versioning_url)
if (http_error(response)) {
  stop("Could not access the Versioning.yml file on GitHub. Please check your internet connection.")
}

# Read the versioning file directly from the URL's content
versioning_data <- read_yaml(text = content(response, "text", encoding = "UTF-8"))

# Look up the details for the user-specified SEF version
version_details <- versioning_data[[config$sef_version]]

# Error handling: Stop if the specified SEF version is not found
if (is.null(version_details)) {
  available_versions <- paste(names(versioning_data), collapse = ", ")
  stop(
    "The specified sef_version '", config$sef_version, "' was not found in the repository.\n",
    "Available versions are: ", available_versions
  )
}

# --- 4. Install and Load correct `useeior` version ---
useeior_tag <- versioning_data[[config$sef_version]][["useeior_tag"]]
useeior_ver <- versioning_data[[config$sef_version]][["useeior_ver"]]
installed_pkg <- installed.packages()
if (!"devtools"%in%installed_pkg[, "Package"]) {
  suppressMessages(install.packages("devtools"))
}
if (!"useeior"%in%installed_pkg[, "Package"]) {
  cli::cli_alert_info("Installing useeior v{useeior_ver} (tag @{useeior_tag}) from GitHub...")
  suppressMessages(devtools::install_github(paste0("USEPA/useeior@", useeior_tag), quiet = TRUE))
}
installed_useeior_ver <- installed_pkg[installed_pkg[, "Package"]=="useeior", "Version"]
if ("useeior"%in%installed_pkg[, "Package"] && useeior_ver!=installed_useeior_ver) {
  cli::cli_alert_warning(c("A new version of useeior (v{useeior_ver}) will be installed for generating SEF {SEF_version}. ",
                           "The useeior v{installed_useeior_ver} you have installed will be overwritten."))
  cli::cli_alert_info("Installing useeior v{useeior_ver} (tag @{useeior_tag}) from GitHub...")
  suppressMessages(devtools::install_github(paste0("USEPA/useeior@", useeior_tag), quiet = TRUE))
}

library(useeior)

cat("Setup complete. Ready to build the model.\n")
```

## 3. Build USEEIO Model

Now, we build the USEEIO model using the `model_name` and the corresponding remote model specification file fetched in the previous step.

```{r 02-build-model, cache=TRUE, echo=FALSE, message=FALSE, warning=FALSE}
# --- 1. Prepare for local model specification file ---
spec_dir <- "spec_files"
# Create the directory if it doesn't exist
if (!dir.exists(spec_dir)) {
  dir.create(spec_dir)
}
# Define the path for the local specification file
# Resolve model_name dynamically from a local mapping that depends on config$sef_version
model_map <- tryCatch(read_yaml("model_spec_mapping.yml"), error = function(e) NULL)
if (!is.null(model_map) && !is.null(model_map$model_name_by_sef_version[[config$sef_version]])) {
  model_name <- model_map$model_name_by_sef_version[[config$sef_version]]
} else {
  # Do not default silently: instruct user to add the mapping explicitly
  msg <- paste0(
    "No model mapping found for sef_version '", config$sef_version, "'.\n",
    "Please add an entry in 'model_spec_mapping.yml' under 'model_name_by_sef_version'.\n",
    "For example:\n",
    "  \"", config$sef_version, "\": \"USEEIOvX.Y.Z-GHG\"\n\n",
    "How to find the correct USEEIO model name: \n",
    " - Check the SEF release notes and/or Versioning.yml in USEPA/supply-chain-factors.\n",
    " - URL: https://raw.githubusercontent.com/USEPA/supply-chain-factors/main/Versioning.yml\n",
    " - Identify the model spec name used for that SEF release (e.g., USEEIOv2.2.22-GHG).\n"
  )
  stop(msg)
}
local_spec_file <- file.path(spec_dir, paste0(model_name, ".yml"))

# --- 2. Download the spec file if it doesn't exist locally ---
if (!file.exists(local_spec_file)) {

  model_spec_url <- paste0(
    "https://raw.githubusercontent.com/USEPA/supply-chain-factors/main/model-specs/",
    model_name,
    ".yml"
  )

  cli::cli_alert_info("Downloading model specification file from: {model_spec_url}")
  response_spec <- GET(model_spec_url)
  
  # Stop with an error if the download fails
  if (http_error(response_spec)) {
    stop("Failed to download the model specification file: ", model_name, ".yml")
  }

  # Write the downloaded content to the local file
  writeBin(content(response_spec, "raw"), local_spec_file)
  cli::cli_alert_success("Model specification saved locally: {local_spec_file}")
} else {
  cli::cli_alert_info("Using existing local model specification: {local_spec_file}")
}

# --- 3. Build the model ---
# We now use the LOCAL file path for the configpaths argument.
# The `cache=TRUE` option will save time on subsequent runs.
cli::cli_alert_info("Building USEEIO model: {model_name} (this can take a minute)...")
model <- buildModel(model_name, configpaths = local_spec_file)
cli::cli_alert_success("Model built.")
```

### Model Specifications

It's important to record the key specifications of the model used for this analysis to ensure reproducibility.

```{r 02a-model-specs, echo=FALSE}
# Extract key specifications from the model object
model_specs <- data.frame(
  Specification = c("USEEIO Supply Chain Emission Factors version", "Model Name", "Input-Output Model's Year", "GHG Data Year", "GHG Data File", "GHG Data File Location", "Model type"),
  Value = c(
    config$sef_version,
    model$specs$Model,
    model$specs$IOYear,
    model$specs$SatelliteTable$GHG$DataYears,
    model$specs$SatelliteTable$GHG$StaticFile,
    model$specs$SatelliteTable$GHG$FileLocation,
    "Commodity"
  )
)

# Print the specifications as a clean table
print(model_specs)
```

## 4. Structural Path Decomposition (SPD)

This is the core of the analysis. We extract matrices from the model and partition them by scope before calculating the tier contributions.

### 4.1 Matrices used (D, A, L, I)

- D: Direct impact vector expanded into a diagonal matrix (per-sector direct GHG intensity)
- A: Direct requirements matrix (technical coefficients)
- L: Leontief inverse $(I - A)^{-1}$ (total requirements)
- I: Identity matrix (used for Economic Tier 1 extraction)

```{r 03a-spd-matrices, echo=FALSE, message=FALSE, warning=FALSE}
# --- 1. Prepare Key Matrices ---
# D: Direct Impact Matrix, diagonally expanded from the vector in model$D
D_matrix <- diag(model$D[1, ])
colnames(D_matrix) <- colnames(model$D)
rownames(D_matrix) <- colnames(model$D)

# I: Identity Matrix (used for Economic Tier 1 calculation)
I_matrix <- diag(nrow(D_matrix))
colnames(I_matrix) <- colnames(D_matrix)
rownames(I_matrix) <- rownames(D_matrix)

# A: Direct Requirements Matrix
A_matrix <- model$A

# A^2: squared Direct Requirements Matrix
A_squared_matrix <- A_matrix %*% A_matrix

# L: Leontief Inverse Matrix
L_matrix <- model$L
```

### 4.2 Partition by Scope

We form Scope 1 and Scope 2 partitions for A and L using the Scope 2 sector list from `config.yml`.

```{r 03b-spd-scope-partitions, echo=FALSE, message=FALSE, warning=FALSE}
# --- 2. Prepare Scope-Specific Matrices ---
# Partition key matrices into Scope 1 and Scope 2 components
A_matrix_scope1 <- A_matrix
A_matrix_scope1[config$scope_2_sectors, ] <- 0

A_matrix_scope2 <- A_matrix
A_matrix_scope2[!(rownames(A_matrix_scope2) %in% config$scope_2_sectors), ] <- 0

# A^2: squared Direct Requirements Matrix
A_squared_matrix_scope2 <- A_squared_matrix
A_squared_matrix_scope2[!(rownames(A_squared_matrix_scope2) %in% config$scope_2_sectors), ] <- 0

# L: Leontief Inverse matrix partitions
L_matrix_scope1 <- L_matrix
L_matrix_scope1[config$scope_2_sectors, ] <- 0

L_matrix_scope2 <- L_matrix
L_matrix_scope2[!(rownames(L_matrix_scope2) %in% config$scope_2_sectors), ] <- 0
```

### 4.3 Compute totals and tiers

We compute total impacts and then allocate them into Economic Tier 1, Economic Tier 2, and Economic Tier 3+ for each Scope.

```{r 03c-spd-totals-and-tiers, echo=FALSE, message=FALSE, warning=FALSE}
# --- 3. Calculate Total Impact for Each Scope ---
# This is used later to calculate the percentage contribution.
N_Total <- D_matrix %*% L_matrix
N_Total_scope1 <- D_matrix %*% L_matrix_scope1
N_Total_scope2 <- D_matrix %*% L_matrix_scope2

# --- 3.1. Economic Tier 1 (Direct Impact: D * I) ---
N_Tier1_scope1 <- D_matrix %*% I_matrix
N_Tier1_scope2 <- D_matrix %*% A_matrix_scope2
N_Tier1 <- N_Tier1_scope1 + N_Tier1_scope2

# --- 3.2. Economic Tier 2 (Direct Suppliers: D * A) ---
N_Tier2_scope1 <- D_matrix %*% A_matrix_scope1
N_Tier2_scope2 <- D_matrix %*% A_squared_matrix_scope2
N_Tier2 <- N_Tier2_scope1 + N_Tier2_scope2

# --- 3.3. Economic Tier 3+ (Residual) ---
N_Tier1_scope1_filtered <- N_Tier1_scope1
N_Tier1_scope1_filtered[rownames(N_Tier1_scope1) %in% config$scope_2_sectors, ] <- 0
N_Tier3plus_scope1 <- N_Total_scope1 - N_Tier1_scope1_filtered - N_Tier2_scope1
N_Tier3plus_scope2 <- N_Total_scope2 - N_Tier1_scope2 - N_Tier2_scope2
N_Tier3plus <- N_Total - N_Tier1 - N_Tier2
```

### Verification Step

Let's verify that our decomposition is correct. The sum of the tiers should equal the total impact matrix.

```{r 04-verification, echo=FALSE, message=FALSE, warning=FALSE}
N_Check <- colSums(N_Tier1 + N_Tier2 + N_Tier3plus)

# Calculate the ratio of the summed tiers to the total impact for each commodity
Ratio_Check <- ifelse(
  abs(model$N) < 1e-12, 
  1,
  N_Check / model$N
)

# Check if all ratio values are close to 1 (within a tolerance for floating-point arithmetic)
verification_passed <- all(abs(Ratio_Check - 1) < 1e-9)
cat(paste("Disaggregation verification successful:", verification_passed, "\n"))
if (!verification_passed) {
  warning("Verification failed! The sum of tiers does not equal the total impact.")
}
```

## 5. Reshape data and build outputs

We now convert matrices into tidy tables by Economic Tier and Scope for each embedded commodity and assemble the final outputs.

### 5.1 Helper for melting matrices

```{r 05a-helper-melt, echo=FALSE, message=FALSE, warning=FALSE}
# --- Helper: reshape matrices to long format with labels ---
melt_and_label <- function(matrix, tier_label, scope_label) {
  # Melt the matrix to long format with explicit row/col identifiers
  # Rows = GHG source sectors, Cols = embedded sector/commodity
  df_long <- reshape2::melt(
    matrix,
    varnames = c("Embedded_Sector_Code_for_GHG_sources", "Disaggregated_Commodity_Code"),
    value.name = "Absolute_Contribution_GHG"
  )

  # Assign displayed Embedded_Sector_Code per requirement:
  # - Scope 1: show under the emitting (source) sector
  # - Scope 2: show under the embedded sector (i.e., the commodity's own sector),
  #            while preserving the electricity (or other source) in _for_GHG_sources
  if (scope_label == "Scope 2") {
    df_long$Embedded_Sector_Code <- df_long$Disaggregated_Commodity_Code
  } else {
    df_long$Embedded_Sector_Code <- df_long$Embedded_Sector_Code_for_GHG_sources
  }

  df_long$Economic_Tier <- tier_label
  df_long$Scope <- scope_label

  # Keep only non-zero entries to reduce file size and improve readability
  df_long <- df_long[df_long$Absolute_Contribution_GHG != 0, ]
  return(df_long)
}
```

### 5.2 Economic Tier 1 and Economic Tier 2 (Scope 1) tables

```{r 05b-build-t1-t2-s1, echo=FALSE, message=FALSE, warning=FALSE}
df_t1_s1 <- melt_and_label(N_Tier1_scope1, "Economic tier 1", "Scope 1")
df_t1_s2 <- melt_and_label(N_Tier1_scope2, "Economic tier 1", "Scope 2")
df_t2_s1 <- melt_and_label(N_Tier2_scope1, "Economic tier 2", "Scope 1")
```

### 5.3 Economic Tier 2 (Scope 2) decomposition via A[j,k]

```{r 05c-build-t2-s2, echo=FALSE, message=FALSE, warning=FALSE}
# M1: contributions from GHG source i to intermediate supplier j (i -> j)
M1 <- D_matrix %*% A_matrix_scope2

# Long form: (Embedded_Sector_Code_for_GHG_sources = i, Embedded_Sector_Code = j, v1)
df_scope2_step1 <- reshape2::melt(
  M1,
  varnames = c("Embedded_Sector_Code_for_GHG_sources", "Embedded_Sector_Code"),
  value.name = "v1"
) %>%
  dplyr::filter(v1 != 0)

# Long form of A: (Embedded_Sector_Code = j, Disaggregated_Commodity_Code = k, a_jk)
df_A_jk <- reshape2::melt(
  A_matrix,
  varnames = c("Embedded_Sector_Code", "Disaggregated_Commodity_Code"),
  value.name = "a_jk"
) %>%
  dplyr::filter(a_jk != 0)

# Join on j to compute i -> j -> k contributions and label as Economic Tier 2, Scope 2
df_t2_s2 <- df_scope2_step1 %>%
  dplyr::inner_join(df_A_jk, by = "Embedded_Sector_Code") %>%
  dplyr::mutate(
    Absolute_Contribution_GHG = v1 * a_jk,
    Economic_Tier = "Economic tier 2",
    Scope = "Scope 2"
  ) %>%
  dplyr::select(
    Disaggregated_Commodity_Code,
    Embedded_Sector_Code,
    Embedded_Sector_Code_for_GHG_sources,
    Economic_Tier, Scope,
    Absolute_Contribution_GHG
  ) %>%
  dplyr::filter(Absolute_Contribution_GHG != 0)
```

### 5.4 Economic Tier 3+ decomposition (Scope 1 and 2)

```{r 05d-build-t3plus, echo=FALSE, message=FALSE, warning=FALSE}
df_t3plus_s1 <- melt_and_label(N_Tier3plus_scope1, "Economic tier 3+", "Scope 1")

# R2plus: captures contribution shares for paths of length >= 2 from j to k
R2plus_matrix <- L_matrix - I_matrix - A_matrix

df_R2plus_jk <- reshape2::melt(
  R2plus_matrix,
  varnames = c("Embedded_Sector_Code", "Disaggregated_Commodity_Code"),
  value.name = "r2plus"
) %>%
  dplyr::filter(r2plus != 0)

# Multiply i->j (v1) by r2plus (j->k via paths length>=2) and label Economic Tier 3+
df_t3plus_s2 <- df_scope2_step1 %>%
  dplyr::inner_join(df_R2plus_jk, by = "Embedded_Sector_Code") %>%
  dplyr::mutate(
    Absolute_Contribution_GHG = v1 * r2plus,
    Economic_Tier = "Economic tier 3+",
    Scope = "Scope 2"
  ) %>%
  dplyr::select(
    Disaggregated_Commodity_Code,
    Embedded_Sector_Code,
    Embedded_Sector_Code_for_GHG_sources,
    Economic_Tier, Scope,
    Absolute_Contribution_GHG
  ) %>%
  dplyr::filter(Absolute_Contribution_GHG != 0)
```

### 5.5a Scope reclassification if scope sector = embedded sector

```{r 05e-scope-reclass, echo=FALSE, message=FALSE, warning=FALSE}
# Reclassify Scope when source and embedded sectors are identical (e.g., the electricity demand of the power sector (T&D losses etc.) is the power sector's scope 1, not 2
df_t1_s2 <- df_t1_s2 %>%
  dplyr::mutate(
    Scope = ifelse(Embedded_Sector_Code == Embedded_Sector_Code_for_GHG_sources, "Scope 1", Scope)
  )

df_t2_s2 <- df_t2_s2 %>%
  dplyr::mutate(
    Scope = ifelse(Embedded_Sector_Code == Embedded_Sector_Code_for_GHG_sources, "Scope 1", Scope)
  )

df_t3plus_s2 <- df_t3plus_s2 %>%
  dplyr::mutate(
    Scope = ifelse(Embedded_Sector_Code == Embedded_Sector_Code_for_GHG_sources, "Scope 1", Scope)
  )
```

### 5.5c Create 221100-upstream (electricity upstream emissions)

```{r 05e-electricity-upstream, echo=FALSE, message=FALSE, warning=FALSE}
# Create a special commodity "221100-upstream_with_losses" representing upstream emissions
# of the electricity sector including losses (all tiers/scopes EXCEPT Tier 1 Scope 1 direct emissions)
electricity_upstream_with_losses <- bind_rows(
  df_t1_s2, df_t2_s1, df_t2_s2, df_t3plus_s1, df_t3plus_s2
) %>%
  # Filter for electricity commodity only
  filter(Disaggregated_Commodity_Code == "221100/US") %>%
  # Clean the /US suffix before filtering
  mutate(
    Disaggregated_Commodity_Code = stringr::str_replace(Disaggregated_Commodity_Code, "/US$", ""),
    Embedded_Sector_Code = stringr::str_replace(Embedded_Sector_Code, "/US$", ""),
    Embedded_Sector_Code_for_GHG_sources = stringr::str_replace(Embedded_Sector_Code_for_GHG_sources, "/US$", "")
  ) %>%
  # Change the commodity code to 221100-upstream_with_losses
  mutate(Disaggregated_Commodity_Code = "221100-upstream_with_losses")

electricity_upstream_without_losses <- bind_rows(
  df_t2_s1, df_t2_s2, df_t3plus_s1, df_t3plus_s2
) %>%
  # Filter for electricity commodity only
  filter(Disaggregated_Commodity_Code == "221100/US") %>%
  # Clean the /US suffix before filtering
  mutate(
    Disaggregated_Commodity_Code = stringr::str_replace(Disaggregated_Commodity_Code, "/US$", ""),
    Embedded_Sector_Code = stringr::str_replace(Embedded_Sector_Code, "/US$", ""),
    Embedded_Sector_Code_for_GHG_sources = stringr::str_replace(Embedded_Sector_Code_for_GHG_sources, "/US$", "")
  ) %>%
  # Change the commodity code to 221100-upstream_without_losses
  mutate(Disaggregated_Commodity_Code = "221100-upstream_without_losses")
```

### 5.6 Build final_table_by_code

```{r 05f-final-table-by-code, echo=FALSE, message=FALSE, warning=FALSE}
# Clean codes, filter Scope 2 commodities from config, aggregate and compute shares
abs_col_name <- paste0(
  "Absolute contribution (kgCO2e/USD_PRO_",
  model$specs$IOYear,
  " spent on embedded commodity)"
)

final_table_by_code <- bind_rows(
    df_t1_s1, df_t1_s2, df_t2_s1, df_t2_s2, df_t3plus_s1, df_t3plus_s2,
    electricity_upstream_with_losses, electricity_upstream_without_losses  # Add the 221100-upstream commodities
  ) %>%
  mutate(
    Disaggregated_Commodity_Code = stringr::str_replace(Disaggregated_Commodity_Code, "/US$", ""),
    Embedded_Sector_Code    = stringr::str_replace(Embedded_Sector_Code, "/US$", ""),
    Embedded_Sector_Code_for_GHG_sources = stringr::str_replace(Embedded_Sector_Code_for_GHG_sources, "/US$", "")
  ) %>%
  group_by(Disaggregated_Commodity_Code) %>%
  mutate(
    Total_Commodity_GHG = sum(Absolute_Contribution_GHG),
    Relative_Contribution = (Absolute_Contribution_GHG / Total_Commodity_GHG)
  ) %>%
  ungroup() %>%
  select(
    Disaggregated_Commodity_Code,
    Embedded_Sector_Code,
    Embedded_Sector_Code_for_GHG_sources,
    Economic_Tier,
    Scope,
    Absolute_Contribution_GHG,
    Relative_Contribution
  ) %>%
  arrange(Disaggregated_Commodity_Code, Economic_Tier, desc(Absolute_Contribution_GHG))
```

```{r 05b-post-aggregation-verification, echo=FALSE, message=FALSE, warning=FALSE}
# Verify that the aggregated contributions sum back to N totals within tolerance
tol <- 1e-9

cat("\n--- Post-aggregation verification steps ---\n")

# Target totals per commodity from N_Total (column sums)
n_total_cols_clean <- stringr::str_replace(colnames(N_Total), "/US$", "")
target_totals <- data.frame(
  Disaggregated_Commodity_Code = n_total_cols_clean,
  N_Total = colSums(N_Total),
  stringsAsFactors = FALSE
)

# Aggregated totals from final_table_by_code (exclude derived commodities like 221100-upstream, and exclude scope 2 commodities)
## Clean scope-2 codes (remove any trailing /US) so we can filter them from verification
scope2_codes_clean <- stringr::str_replace(config$scope_2_sectors, "/US$", "")

agg_totals <- final_table_by_code %>%
  dplyr::group_by(Disaggregated_Commodity_Code) %>%
  dplyr::summarise(Sum_From_Table = sum(Absolute_Contribution_GHG), .groups = "drop")

merged_check <- agg_totals %>% dplyr::inner_join(target_totals, by = "Disaggregated_Commodity_Code") %>%
  dplyr::mutate(
    abs_diff = abs(Sum_From_Table - N_Total),
    rel_diff = ifelse(abs(N_Total) < tol, abs_diff, abs_diff / pmax(abs(N_Total), tol)),
    pass = abs_diff <= tol | rel_diff <= tol
  )

overall_pass <- all(merged_check$pass)
cat("Step 1: Compare per-commodity sums from table vs. N_Total column sums\n")
cat(" - Commodities checked:", nrow(merged_check), "\n")
cat(" - PASS:", overall_pass, "\n")

if (!overall_pass) {
  # Determine columns (commodities) to include in secondary checks (those present in the table)
  keep_cols <- which(n_total_cols_clean %in% merged_check$Disaggregated_Commodity_Code)

  # Secondary checks by Scope totals across all kept commodities (exclude derived commodities)
  scope1_target_total <- sum(N_Total_scope1[, keep_cols, drop = FALSE])
  scope2_target_total <- sum(N_Total_scope2[, keep_cols, drop = FALSE])
  scope1_table_total  <- sum(final_table_by_code$Absolute_Contribution_GHG[
    final_table_by_code$Scope == "Scope 1" & 
    !final_table_by_code$Disaggregated_Commodity_Code %in% c("221100-upstream")
  ]) 
  scope2_table_total  <- sum(final_table_by_code$Absolute_Contribution_GHG[
    final_table_by_code$Scope == "Scope 2" & 
    !final_table_by_code$Disaggregated_Commodity_Code %in% c("221100-upstream")
  ]) 

  cat("Step 2: Check totals by Scope across kept commodities\n")
  cat(sprintf(" - Scope 1: table=%.12g target=%.12g abs_diff=%.12g\n", scope1_table_total, scope1_target_total, abs(scope1_table_total - scope1_target_total)))
  cat(sprintf(" - Scope 2: table=%.12g target=%.12g abs_diff=%.12g\n", scope2_table_total, scope2_target_total, abs(scope2_table_total - scope2_target_total)))

  # Tertiary checks by Tier x Scope across all kept commodities
  t1s1_target <- sum(N_Tier1_scope1[, keep_cols, drop = FALSE])
  t1s2_target <- sum(N_Tier1_scope2[, keep_cols, drop = FALSE])
  t2s1_target <- sum(N_Tier2_scope1[, keep_cols, drop = FALSE])
  t2s2_target <- sum(N_Tier2_scope2[, keep_cols, drop = FALSE])
  t3s1_target <- sum(N_Tier3plus_scope1[, keep_cols, drop = FALSE])
  t3s2_target <- sum(N_Tier3plus_scope2[, keep_cols, drop = FALSE])

  t1s1_table <- sum(final_table_by_code$Absolute_Contribution_GHG[
    final_table_by_code$Economic_Tier == "Economic tier 1" & 
    final_table_by_code$Scope == "Scope 1" &
    !final_table_by_code$Disaggregated_Commodity_Code %in% c("221100-upstream")
  ]) 
  t1s2_table <- sum(final_table_by_code$Absolute_Contribution_GHG[
    final_table_by_code$Economic_Tier == "Economic tier 1" & 
    final_table_by_code$Scope == "Scope 2" &
    !final_table_by_code$Disaggregated_Commodity_Code %in% c("221100-upstream")
  ]) 
  t2s1_table <- sum(final_table_by_code$Absolute_Contribution_GHG[
    final_table_by_code$Economic_Tier == "Economic tier 2" & 
    final_table_by_code$Scope == "Scope 1" &
    !final_table_by_code$Disaggregated_Commodity_Code %in% c("221100-upstream")
  ]) 
  t2s2_table <- sum(final_table_by_code$Absolute_Contribution_GHG[
    final_table_by_code$Economic_Tier == "Economic tier 2" & 
    final_table_by_code$Scope == "Scope 2" &
    !final_table_by_code$Disaggregated_Commodity_Code %in% c("221100-upstream")
  ]) 
  t3s1_table <- sum(final_table_by_code$Absolute_Contribution_GHG[
    final_table_by_code$Economic_Tier == "Economic tier 3+" & 
    final_table_by_code$Scope == "Scope 1" &
    !final_table_by_code$Disaggregated_Commodity_Code %in% c("221100-upstream")
  ]) 
  t3s2_table <- sum(final_table_by_code$Absolute_Contribution_GHG[
    final_table_by_code$Economic_Tier == "Economic tier 3+" & 
    final_table_by_code$Scope == "Scope 2" &
    !final_table_by_code$Disaggregated_Commodity_Code %in% c("221100-upstream")
  ]) 

  cat("Step 3: Check totals by Tier x Scope across kept commodities\n")
  cat(sprintf(" - Tier1 Scope1: table=%.12g target=%.12g abs_diff=%.12g\n", t1s1_table, t1s1_target, abs(t1s1_table - t1s1_target)))
  cat(sprintf(" - Tier1 Scope2: table=%.12g target=%.12g abs_diff=%.12g\n", t1s2_table, t1s2_target, abs(t1s2_table - t1s2_target)))
  cat(sprintf(" - Tier2 Scope1: table=%.12g target=%.12g abs_diff=%.12g\n", t2s1_table, t2s1_target, abs(t2s1_table - t2s1_target)))
  cat(sprintf(" - Tier2 Scope2: table=%.12g target=%.12g abs_diff=%.12g\n", t2s2_table, t2s2_target, abs(t2s2_table - t2s2_target)))
  cat(sprintf(" - Tier3+ Scope1: table=%.12g target=%.12g abs_diff=%.12g\n", t3s1_table, t3s1_target, abs(t3s1_table - t3s1_target)))
  cat(sprintf(" - Tier3+ Scope2: table=%.12g target=%.12g abs_diff=%.12g\n", t3s2_table, t3s2_target, abs(t3s2_table - t3s2_target)))

  # Show a few worst commodity mismatches to aid debugging
  worst <- merged_check %>% dplyr::arrange(dplyr::desc(abs_diff)) %>% utils::head(5)
  cat("Top 5 commodity mismatches (table vs. N_Total):\n")
  print(worst)
}

# Additional check: Verify Scope 2 sectors match N_Total
if (length(scope2_codes_clean) > 0) {
  cat("\nStep 4: Verify Scope 2 sectors match N_Total\n")
  scope2_totals <- final_table_by_code %>%
    dplyr::filter(Disaggregated_Commodity_Code %in% scope2_codes_clean) %>%
    dplyr::group_by(Disaggregated_Commodity_Code) %>%
    dplyr::summarise(Sum_From_Table = sum(Absolute_Contribution_GHG), .groups = "drop") %>%
    dplyr::inner_join(target_totals, by = "Disaggregated_Commodity_Code") %>%
    dplyr::mutate(
      abs_diff = abs(Sum_From_Table - N_Total),
      rel_diff = ifelse(abs(N_Total) < tol, abs_diff, abs_diff / pmax(abs(N_Total), tol)),
      pass = abs_diff <= tol | rel_diff <= tol
    )
  
  scope2_pass <- all(scope2_totals$pass)
  cat(" - Scope 2 sectors checked:", nrow(scope2_totals), "\n")
  cat(" - PASS:", scope2_pass, "\n")
  
  if (!scope2_pass) {
    cat("Scope 2 sector mismatches:\n")
    print(scope2_totals %>% dplyr::filter(!pass))
  }
}

cat("--- End verification ---\n\n")
```

```{r supporting-tables, echo=FALSE, message=FALSE, warning=FALSE}
# --- 4. Create Supporting Tables for Excel Export ---

# Lightweight progress logger
log_step <- function(msg){
  cat(sprintf("[%s] %s\n", format(Sys.time(), "%H:%M:%S"), msg))
  utils::flush.console()
}

log_step("Building supporting tables (names, electricity-only views, sector classification)...")

# Table with Names instead of Codes
industry_mapping <- model$Industries %>% select(Code, Name)
commodity_mapping <- model$Commodities %>% select(Code, Name)

final_table_by_name <- final_table_by_code %>%
  left_join(commodity_mapping, by = c("Disaggregated_Commodity_Code" = "Code")) %>%
  rename(Disaggregated_Commodity_Name = Name) %>%
  mutate(
    Disaggregated_Commodity_Name = dplyr::case_when(
      Disaggregated_Commodity_Code == "221100-upstream_with_losses" ~ "Electricity (upstream with losses)",
      Disaggregated_Commodity_Code == "221100-upstream_without_losses" ~ "Electricity (upstream without losses)",
      TRUE ~ Disaggregated_Commodity_Name
    )
  ) %>%
  left_join(industry_mapping, by = c("Embedded_Sector_Code" = "Code")) %>%
  rename(Embedded_Sector_Name = Name) %>%
  left_join(industry_mapping, by = c("Embedded_Sector_Code_for_GHG_sources" = "Code")) %>%
  rename(Embedded_Sector_Name_for_GHG_sources = Name) %>%
  select(
    Disaggregated_Commodity_Code,
    Disaggregated_Commodity_Name,
    Embedded_Sector_Code,
    Embedded_Sector_Code_for_GHG_sources,
    Embedded_Sector_Name,
    Embedded_Sector_Name_for_GHG_sources,
    Economic_Tier,
    Scope,
    Absolute_Contribution_GHG,
    Relative_Contribution
  )

log_step(sprintf("final_table_by_name ready (%s rows)", format(nrow(final_table_by_name), big.mark=",")))

# Electricity-only filtered views (by GHG source sector)
# Use fixed code for electricity as requested
electricity_codes <- c("221100")
electric_only_by_code <- final_table_by_code %>%
  dplyr::filter(Embedded_Sector_Code_for_GHG_sources %in% electricity_codes)
electric_only_by_name <- final_table_by_name %>%
  dplyr::filter(Embedded_Sector_Code_for_GHG_sources %in% electricity_codes)

log_step(sprintf("Electricity-only tables ready (code rows: %s, name rows: %s)",
                 format(nrow(electric_only_by_code), big.mark=","),
                 format(nrow(electric_only_by_name), big.mark=",")))

# Table with Industry Details
sector_classification <- model$Industries %>%
  select(Code, `Industry name` = Name) %>%
  left_join(
    model$Commodities %>% select(Code, `Commodity name` = Name, Category, Subcategory, Description),
    by = "Code"
  ) %>%
  # Separate Category into Code and Name
  tidyr::separate(Category, into = c("Category Code", "Category Name"), sep = ": ", extra = "merge", fill = "right", remove = TRUE) %>%
  # Separate Subcategory into Code and Name
  tidyr::separate(Subcategory, into = c("Subcategory Code", "Subcategory Name"), sep = ": ", extra = "merge", fill = "right", remove = TRUE) %>%
  mutate(
    # Use a single regex to remove multiple possible prefixes
    Description = stringr::str_remove(Description, "^(BEA Code & Name is '?\\w+:.*?'?\\.\\s*)")
    ) %>%
  # Select and reorder the final columns for the Excel sheet
  select(
    `Category Code`,
    `Category Name`,
    `Subcategory Code`,
    `Subcategory Name`,
    `Sector code` = Code,
    `Sector name` = `Industry name`,
    `Commodity name`,
    Description
  )

# Append derived electricity commodity rows
electricity_derived_rows <- tibble::tibble(
  `Category Code`    = "22",
  `Category Name`    = "Utilities",
  `Subcategory Code` = "221",
  `Subcategory Name` = "Utilities",
  `Sector code`      = c("221100-upstream_with_losses", "221100-upstream_without_losses"),
  `Sector name`      = c("Electricity (upstream with losses)", "Electricity (upstream without losses)"),
  `Commodity name`   = c("Electricity (upstream with losses)", "Electricity (upstream without losses)"),
  Description        = c(
    "Derived commodity: upstream GHG emissions of the electricity sector including T&D losses (all tiers/scopes except Tier 1 Scope 1 direct emissions).",
    "Derived commodity: upstream GHG emissions of the electricity sector excluding T&D losses (Tier 2+ only)."
  )
)
sector_classification <- bind_rows(sector_classification, electricity_derived_rows)

log_step(sprintf("sector_classification ready (%s rows)", format(nrow(sector_classification), big.mark=",")))

```

```{r write-tables, echo=FALSE, message=FALSE, warning=FALSE}
# --- 5. Write all tables to a single Excel file ---
# Create a clean base filename from the model specs and ensure an outputs directory exists
outputs_dir <- "outputs"
if (!dir.exists(outputs_dir)) dir.create(outputs_dir, recursive = TRUE)

# Create format subdirectories
csv_dir <- file.path(outputs_dir, "csv")
parquet_dir <- file.path(outputs_dir, "parquet")
jsonld_dir <- file.path(outputs_dir, "jsonld")
for (d in c(csv_dir, parquet_dir, jsonld_dir)) {
  if (!dir.exists(d)) dir.create(d, recursive = TRUE)
}

file_basename <- paste(
  "SEF",
  config$sef_version,
  "_disaggregation_factors_",
  paste0("GHG", model_specs[model_specs$Specification == "GHG Data Year", "Value"]),
  sep = "_",
  paste0("IO", model_specs[model_specs$Specification == "Input-Output Model's Year", "Value"])
)

# Excel in main outputs directory
excel_filename <- file.path(outputs_dir, paste0(file_basename, ".xlsx"))

# CSV files in csv/ subdirectory
csv_filename   <- file.path(csv_dir, paste0(file_basename, ".csv"))
csv_elec_filename <- file.path(csv_dir, paste0(file_basename, "_electricity_only.csv"))
csv_sc_filename   <- file.path(csv_dir, "sector_classification.csv")
csv_light_filename <- file.path(csv_dir, paste0(file_basename, "_light.csv"))

# Read export toggles from config (default all to TRUE if not specified)
export_excel <- if(!is.null(config$export_options$excel)) config$export_options$excel else TRUE
export_csv_contributions <- if(!is.null(config$export_options$csv_contributions)) config$export_options$csv_contributions else TRUE
export_csv_lightweight <- if(!is.null(config$export_options$csv_lightweight)) config$export_options$csv_lightweight else TRUE
export_csv_electricity <- if(!is.null(config$export_options$csv_electricity_only)) config$export_options$csv_electricity_only else TRUE
export_csv_sector_class <- if(!is.null(config$export_options$csv_sector_classification)) config$export_options$csv_sector_classification else TRUE

if (export_excel) {
  wb <- createWorkbook()
  log_step("Creating Excel workbook and writing sheets...")

# --- Create and Add Author Info Tab (as the first sheet) ---
author_info <- data.frame(
  Field = c(
    "Author",
    "Organization",
    "Website",
    "Contact",
    "Open-source repository",
    "Q&A + Discussion",
    "Data License",
    "License URL",
    "",
    "Acknowledgments",
    "USEEIO model & useeior package",
    "Supply Chain Emission Factors",
    "Upstream data license"
  ),
  Value = c(
    "Damien Lieber",
    "DecarbNexus LLC",
    "decarbnexus.com",
    "contact@decarbnexus.com",
    "https://github.com/DecarbNexus/useeio_sectors_disaggregation",
    "https://github.com/DecarbNexus/useeio_sectors_disaggregation/discussions",
    "CC BY 4.0 (see LICENSE-DATA.md)",
    "https://creativecommons.org/licenses/by/4.0/",
    "",
    "This project builds on open-source work by the U.S. EPA (see below)",
    "https://github.com/USEPA/useeior (MIT License, Copyright U.S. EPA)",
    "https://github.com/USEPA/supply-chain-factors (MIT License, Copyright U.S. EPA)",
    "Both upstream projects are released under the MIT License. See their repositories for full license text."
  ),
  stringsAsFactors = FALSE
)
addWorksheet(wb, "Author_Info")
writeData(wb, "Author_Info", author_info)
log_step("Wrote sheet: Author_Info")

addWorksheet(wb, "Model_Specs")
writeData(wb, "Model_Specs", model_specs)
log_step("Wrote sheet: Model_Specs")

addWorksheet(wb, "Contributions_by_Name")
writeData(wb, "Contributions_by_Name", final_table_by_name %>%
  rename(!!abs_col_name := Absolute_Contribution_GHG))
log_step(sprintf("Wrote sheet: Contributions_by_Name (%s rows)", format(nrow(final_table_by_name), big.mark=",")))

addWorksheet(wb, "Contributions_by_Code")
writeData(wb, "Contributions_by_Code", final_table_by_code %>%
  rename(!!abs_col_name := Absolute_Contribution_GHG))
log_step(sprintf("Wrote sheet: Contributions_by_Code (%s rows)", format(nrow(final_table_by_code), big.mark=",")))

addWorksheet(wb, "Electricity_only")
writeData(wb, "Electricity_only", electric_only_by_name %>%
  rename(!!abs_col_name := Absolute_Contribution_GHG))
log_step(sprintf("Wrote sheet: Electricity_only (%s rows)", format(nrow(electric_only_by_name), big.mark=",")))

addWorksheet(wb, "sector_classification")
writeData(wb, "sector_classification", sector_classification)
log_step(sprintf("Wrote sheet: sector_classification (%s rows)", format(nrow(sector_classification), big.mark=",")))

addWorksheet(wb, "NAICS_to_USEEIO_crosswalk")
writeData(wb, "NAICS_to_USEEIO_crosswalk", model$crosswalk)
log_step(sprintf("Wrote sheet: NAICS_to_USEEIO_crosswalk (%s rows)", format(nrow(model$crosswalk), big.mark=",")))

# Matrices: Direct impact (D), Direct requirements (A), Leontief inverse (L)
addWorksheet(wb, "Direct GHG impacts (d_GHG)")
writeData(wb, "Direct GHG impacts (d_GHG)", t(model$D), rowNames = TRUE)
log_step("Wrote sheet: Direct GHG impacts (d_GHG)")

#addWorksheet(wb, "Diagonalized d_GHG (D_GHG)")
#writeData(wb, "Diagonalized d_GHG (D_GHG)", D_matrix, rowNames = TRUE)
#log_step("Wrote sheet: Diagonalized d_GHG (D_GHG)")

addWorksheet(wb, "Direct requirements matrix (A)")
writeData(wb, "Direct requirements matrix (A)", A_matrix, rowNames = TRUE)
log_step("Wrote sheet: Direct requirements matrix (A)")

addWorksheet(wb, "Leontief inverse (L)")
writeData(wb, "Leontief inverse (L)", L_matrix, rowNames = TRUE)
log_step("Wrote sheet: Leontief inverse (L)")

addWorksheet(wb, "Total impacts from useeior (N)")
N_export <- t(model$N)
rownames(N_export) <- stringr::str_replace(rownames(N_export), "/US$", "")
writeData(wb, "Total impacts from useeior (N)", N_export, rowNames = TRUE)
log_step("Wrote sheet: Total impacts from useeior (N)")

  saveWorkbook(wb, excel_filename, overwrite = TRUE)
  log_step(sprintf("Saved Excel: %s", excel_filename))
} else {
  log_step("Skipped Excel export (disabled in config.yml)")
}

# Write a short data license notice into outputs/
data_license_text <- c(
  "Data License: CC BY 4.0",
  "Full license: https://creativecommons.org/licenses/by/4.0/",
  "If you use these datasets, cite DecarbNexus LLC and acknowledge USEPA USEEIO/SEF sources:",
  " - USEEIO: https://github.com/USEPA/useeior",
  " - Supply Chain Emission Factors: https://github.com/USEPA/supply-chain-factors"
)
writeLines(data_license_text, file.path(outputs_dir, "DATA_LICENSE.txt"))
log_step("Wrote file: outputs/DATA_LICENSE.txt")

# Main contributions CSV export
if (export_csv_contributions) {
  write.csv(final_table_by_code %>%
    rename(!!abs_col_name := Absolute_Contribution_GHG), csv_filename, row.names = FALSE, na = "")
  log_step(sprintf("Wrote CSV: %s", csv_filename))
} else {
  log_step("Skipped main contributions CSV (disabled in config.yml)")
}

# Electricity-only CSV export
if (export_csv_electricity) {
  write.csv(electric_only_by_code %>%
    rename(!!abs_col_name := Absolute_Contribution_GHG), csv_elec_filename, row.names = FALSE, na = "")
  log_step(sprintf("Wrote CSV: %s", csv_elec_filename))
} else {
  log_step("Skipped electricity-only CSV (disabled in config.yml)")
}

# Export sector classification as CSV
if (export_csv_sector_class) {
  write.csv(
    sector_classification,
    csv_sc_filename,
    row.names = FALSE,
    na = ""
  )
  log_step(sprintf("Wrote CSV: %s", csv_sc_filename))
} else {
  log_step("Skipped sector classification CSV (disabled in config.yml)")
}

# Lightweight CSV export: commodity > embedded sector > tier > scope > relative_contribution
if (export_csv_lightweight) {
  csv_light_data <- final_table_by_code %>%
    group_by(
      Disaggregated_Commodity_Code,
      Embedded_Sector_Code,
      Economic_Tier,
      Scope
    ) %>%
    summarise(
      Relative_Contribution = sum(Relative_Contribution),
      .groups = "drop"
    )
  
  write.csv(
    csv_light_data,
    csv_light_filename,
    row.names = FALSE,
    na = ""
  )
  log_step(sprintf("Wrote lightweight CSV: %s", csv_light_filename))
} else {
  log_step("Skipped lightweight CSV (disabled in config.yml)")
}

cat("\n--- Export Summary ---\n")
if (export_excel) cat("Saved Excel:", excel_filename, "\n")
if (export_csv_contributions) cat("Saved Contributions CSV:", csv_filename, "\n")
if (export_csv_lightweight) cat("Saved Lightweight CSV:", csv_light_filename, "\n")
if (export_csv_electricity) cat("Saved Electricity-only CSV:", csv_elec_filename, "\n")
if (export_csv_sector_class) cat("Saved Sector Classification CSV:", csv_sc_filename, "\n")

# --- 6. Structured outputs for data science and knowledge graphs ---
log_step("Writing structured outputs (Parquet, JSON-LD)...")

# Read export toggles from config (default all to TRUE if not specified)
export_parquet <- if(!is.null(config$export_options$parquet)) config$export_options$parquet else TRUE
export_jsonld <- if(!is.null(config$export_options$jsonld)) config$export_options$jsonld else TRUE

# 6.1 Parquet: optimized columnar format for data science
if (export_parquet) {
  parquet_start <- Sys.time()
  parquet_filename <- file.path(parquet_dir, paste0(file_basename, ".parquet"))
  parquet_sc_filename <- file.path(parquet_dir, "sector_classification.parquet")
  arrow::write_parquet(
    final_table_by_code %>% rename(!!abs_col_name := Absolute_Contribution_GHG),
    parquet_filename,
    compression = "snappy"
  )
  arrow::write_parquet(sector_classification, parquet_sc_filename, compression = "snappy")
  parquet_elapsed <- round(as.numeric(difftime(Sys.time(), parquet_start, units = "secs")), 2)
  log_step(sprintf("Wrote Parquet (2 files) in %.2f seconds", parquet_elapsed))
} else {
  log_step("Skipped Parquet export (disabled in config.yml)")
}

# 6.2 JSON-LD: RDF-ready with @context for knowledge graph ingestion and D3.js visualization
if (export_jsonld) {
  jsonld_start <- Sys.time()
  log_step("Building hierarchical JSON-LD structures (full + lightweight)...")
  
  # === FULL VERSION: All detail preserved ===
  # Structure: commodity → tier → sector → scope (with ghg_source, scope type, contributions)
  log_step("Building full JSON-LD (commodity > tier > sector > scope)...")
  json_data_full <- final_table_by_code %>%
    transmute(
      commodity_code = Disaggregated_Commodity_Code,
      tier = Economic_Tier,
      sector_code = Embedded_Sector_Code,
      scope = Scope,
      ghg_source_code = Embedded_Sector_Code_for_GHG_sources,
      absolute_contribution = Absolute_Contribution_GHG,
      relative_contribution = Relative_Contribution
    )
  
  # Build nested hierarchy: commodity -> tier -> sector -> scopes
  nested_json_full <- json_data_full %>%
    group_by(commodity_code, tier, sector_code) %>%
    summarise(
      scopes = list(tibble(
        scope = scope,
        ghg_source_code = ghg_source_code,
        absolute_contribution = absolute_contribution,
        relative_contribution = relative_contribution
      )),
      .groups = "drop"
    ) %>%
    group_by(commodity_code, tier) %>%
    summarise(
      sectors = list(tibble(
        sector_code = sector_code,
        scopes = scopes
      )),
      .groups = "drop"
    ) %>%
    group_by(commodity_code) %>%
    summarise(
      tiers = list(tibble(
        tier = tier,
        sectors = sectors
      )),
      .groups = "drop"
    )
  
  # === LIGHTWEIGHT VERSION: Aggregated for simple viz ===
  # Structure: commodity → tier → sector (with name, scope, relative_contribution)
  log_step("Building lightweight JSON-LD (commodity > tier > sector > scope)...")
  json_data_light <- final_table_by_code %>%
    left_join(
      industry_mapping,
      by = c("Embedded_Sector_Code" = "Code")
    ) %>%
    group_by(
      Disaggregated_Commodity_Code,
      Economic_Tier,
      Embedded_Sector_Code,
      Name,
      Scope
    ) %>%
    summarise(
      relative_contribution = sum(Relative_Contribution),
      .groups = "drop"
    ) %>%
    transmute(
      commodity_code = Disaggregated_Commodity_Code,
      tier = Economic_Tier,
      embedded_sector_code = Embedded_Sector_Code,
      embedded_sector_name = Name,
      scope = Scope,
      relative_contribution = relative_contribution
    )
  
  nested_json_light <- json_data_light %>%
    group_by(commodity_code, tier) %>%
    summarise(
      sectors = list(tibble(
        embedded_sector_code = embedded_sector_code,
        embedded_sector_name = embedded_sector_name,
        scope = scope,
        relative_contribution = relative_contribution
      )),
      .groups = "drop"
    ) %>%
    group_by(commodity_code) %>%
    summarise(
      tiers = list(tibble(
        tier = tier,
        sectors = sectors
      )),
      .groups = "drop"
    )
  
  # Build nested sector classification: category -> subcategory -> sectors
  sc_json_nested <- sector_classification %>%
    group_by(`Category Code`, `Category Name`, `Subcategory Code`, `Subcategory Name`) %>%
    summarise(
      sectors = list(tibble(
        sector_code = `Sector code`,
        sector_name = `Sector name`,
        commodity_name = `Commodity name`,
        description = Description
      )),
      .groups = "drop"
    ) %>%
    group_by(`Category Code`, `Category Name`) %>%
    summarise(
      subcategories = list(tibble(
        subcategory_code = `Subcategory Code`,
        subcategory_name = `Subcategory Name`,
        sectors = sectors
      )),
      .groups = "drop"
    ) %>%
    transmute(
      category_code = `Category Code`,
      category_name = `Category Name`,
      subcategories = subcategories
    )
  
  # Sector classification JSON-LD with hierarchical structure
  sc_jsonld_context <- list(
    "@vocab" = "https://decarbnexus.com/vocab/useeio-classification#",
    "schema" = "http://schema.org/",
    "category_code" = list("@id" = "categoryCode", "@type" = "schema:Text"),
    "category_name" = list("@id" = "categoryName", "@type" = "schema:Text"),
    "subcategory_code" = list("@id" = "subcategoryCode", "@type" = "schema:Text"),
    "subcategory_name" = list("@id" = "subcategoryName", "@type" = "schema:Text"),
    "sector_code" = list("@id" = "sectorCode", "@type" = "@id"),
    "sector_name" = list("@id" = "sectorName", "@type" = "schema:Text"),
    "commodity_name" = list("@id" = "commodityName", "@type" = "schema:Text"),
    "description" = list("@id" = "schema:description", "@type" = "schema:Text"),
    "subcategories" = "hasSubcategory",
    "sectors" = "hasSector"
  )
  
  # Define vocabularies for both full and lightweight JSON-LD
  
  # Full version vocabulary (includes sector and scope)
  jsonld_context_full <- list(
    "@vocab" = "https://decarbnexus.com/vocab/useeio-disaggregation#",
    "schema" = "http://schema.org/",
    "useeio" = "https://www.epa.gov/land-research/us-environmentally-extended-input-output-useeio-technical-content#",
    "commodity_code" = list("@id" = "useeio:commodity", "@type" = "@id"),
    "sector_code" = list("@id" = "useeio:sector", "@type" = "@id"),
    "ghg_source_code" = list("@id" = "ghgSource", "@type" = "@id"),
    "tier" = "economicTier",
    "scope" = "ghgScope",
    "absolute_contribution" = list("@id" = "absoluteContribution", "@type" = "schema:Float"),
    "relative_contribution" = list("@id" = "relativeContribution", "@type" = "schema:Float"),
    "tiers" = "hasTier",
    "sectors" = "hasSector",
    "scopes" = "hasScope"
  )
  
  # Lightweight version vocabulary (with sector name and scope)
  jsonld_context_light <- list(
    "@vocab" = "https://decarbnexus.com/vocab/useeio-disaggregation#",
    "schema" = "http://schema.org/",
    "useeio" = "https://www.epa.gov/land-research/us-environmentally-extended-input-output-useeio-technical-content#",
    "commodity_code" = list("@id" = "useeio:commodity", "@type" = "@id"),
    "embedded_sector_code" = list("@id" = "useeio:sector", "@type" = "@id"),
    "embedded_sector_name" = list("@id" = "sectorName", "@type" = "schema:Text"),
    "tier" = "economicTier",
    "scope" = "ghgScope",
    "relative_contribution" = list("@id" = "relativeContribution", "@type" = "schema:Float"),
    "tiers" = "hasTier",
    "sectors" = "hasSector"
  )
  
  jsonld_output_full <- list(
    "@context" = jsonld_context_full,
    "@graph" = nested_json_full
  )
  
  jsonld_output_light <- list(
    "@context" = jsonld_context_light,
    "@graph" = nested_json_light
  )
  
  sc_jsonld_output <- list(
    "@context" = sc_jsonld_context,
    "@graph" = sc_json_nested
  )
  
  log_step("Writing JSON-LD files...")
  
  jsonld_filename <- file.path(jsonld_dir, paste0(file_basename, ".jsonld"))
  jsonld_light_filename <- file.path(jsonld_dir, paste0(file_basename, "_light.jsonld"))
  jsonld_sc_filename <- file.path(jsonld_dir, "sector_classification.jsonld")
  
  # Write full JSON-LD
  jsonlite::write_json(
    jsonld_output_full,
    jsonld_filename,
    pretty = TRUE,
    auto_unbox = TRUE,
    digits = 6
  )
  
  # Write lightweight JSON-LD
  jsonlite::write_json(
    jsonld_output_light,
    jsonld_light_filename,
    pretty = TRUE,
    auto_unbox = TRUE,
    digits = 6
  )
  
  # Write sector classification
  jsonlite::write_json(
    sc_jsonld_output,
    jsonld_sc_filename,
    pretty = TRUE,
    auto_unbox = TRUE
  )
  
  jsonld_elapsed <- round(as.numeric(difftime(Sys.time(), jsonld_start, units = "secs")), 2)
  log_step(sprintf("Wrote JSON-LD (3 files: full, lightweight, classification) in %.2f seconds", jsonld_elapsed))
} else {
  log_step("Skipped JSON-LD export (disabled in config.yml)")
}

# Summary of structured outputs timing
if (export_parquet || export_jsonld) {
  total_structured_elapsed <- 
    (if(export_parquet) parquet_elapsed else 0) +
    (if(export_jsonld) jsonld_elapsed else 0)
  log_step(sprintf("All structured outputs complete in %.2f seconds total", total_structured_elapsed))
}

if (export_parquet || export_jsonld) {
  cat("\n--- Structured Outputs ---\n")
  if (export_parquet) {
    cat("Saved Parquet (data):", parquet_filename, "\n")
    cat("Saved Parquet (class):", parquet_sc_filename, "\n")
  }
  if (export_jsonld) {
    cat("Saved JSON-LD (full):", jsonld_filename, "\n")
    cat("Saved JSON-LD (light):", jsonld_light_filename, "\n")
    cat("Saved JSON-LD (class):", jsonld_sc_filename, "\n")
  }
}
```

## 6. Preview of Final Results

Here is a preview of the top 5 rows from the final contribution factor table (with codes). The full results, including the version with names, are in the Excel file.

```{r 06-preview-results, echo=FALSE, message=FALSE, warning=FALSE}
# Preview with helpful context and an attempt to show all columns
cat("\n--- Preview of Final Results ---\n")
cat("Showing top 5 rows. Attempting to display all columns; if your console truncates, a compact view follows.\n\n")

# Try printing all columns (tibble-friendly); fall back to base data.frame
printed_full <- FALSE
try({
  if (requireNamespace("tibble", quietly = TRUE)) {
    print(tibble::as_tibble(final_table_by_name) |> utils::head(5), n = 5, width = Inf)
    printed_full <- TRUE
  }
}, silent = TRUE)

if (!printed_full) {
  print(utils::head(as.data.frame(final_table_by_name), 5), row.names = FALSE)
}

# Compact view with key columns (for narrow terminals)
key_cols <- c("Disaggregated_Commodity_Code", "Embedded_Sector_Code", "Economic_Tier", "Scope", "Relative_Contribution")
missing_cols <- setdiff(key_cols, names(final_table_by_name))
if (length(missing_cols) == 0) {
  cat("\nCompact view (key columns):\n")
  print(utils::head(final_table_by_name[ , key_cols, drop = FALSE], 5), row.names = FALSE)
} else {
  cat("\nNote: Some key columns were not found:", paste(missing_cols, collapse = ", "), "\n")
}

cat("--- End Preview ---\n\n")
```